Testing the Relationship between Phenomenological Control related to Illusion Sensitivity

### 2025-02-28 paper/R1/manuscript.qmd
Edits to manuscript
Un-bolded old comments, checked everything that said "addressed" and moved to new R1 folder.

Visual Illusions are an interesting type of stimuli highlighting the ease with which our phenomenological conscious experience can become dissociated from physical reality. Their robust and reliable effect makes them useful stimuli to explore how perception is constructed and shaped, and several theoretical models have been put forth to explain how they work. In particular, illusions have been recently reframed using a predictive coding account of perception [@notredame2014] in which the brain optimally combines, using some flavour of Bayesian inference, perceptual inputs with prior knowledge to make sense of ambiguous environments [@friston2010].

@@

However, the exact nature of this interindividual variability and its potential origin remains unclear. The somewhat mixed evidence in the literature regarding its generalizability and strength could be related to the variety of the paradigms used and the type of processes being mobilised [@makowski2021]. Indeed, traditional methods frequently focus on participant's experience by prompting them to assess the difference between two identical targets, estimate the target's physical properties, or adjust the targets to match a reference stimulus [@todorovic2020]. Relying on metacognitive judgments about one's subjective experiences adds an additional layer to the measure that might not be desired when attempting to measure illusion sensibility. Moreover, paradigms often face challenges in diversifying the illusory effects (i.e., using multiple stimuli to experimentally manipulate the strength of the illusion) and the illusion types (i.e., using various illusions, such as Müller-Lyer, Ebbinghaus, Delboeuf which might rely on a different admixture of mechanisms), hindering the potential of obtaining a comprehensive, valid, and reliable measure of illusion sensitivity.

@@

To encourage the empirical exploration of our ability and tendency to alter our phenomenological experience and further accelerate investigations away from the hypnotic context, @lush2021 adapted the Sussex-Waterloo Scale of Hypnotisability [SWASH, @lush2018] by removing all its references to hypnosis, to measure trait phenomenological control. This newly developed phenomenological control scale (PCS) consists of 10 imaginative suggestions followed by subjective ratings for each suggestion on a 6-point Likert scale and has demonstrated validity in online experiments [@lush2022].

@@

The experiment’s setup follows of the born-open principle [@deleeuw2023]. The online experiment, implemented entirely using JsPsych [@de2015jspsych], has its code stored on GitHub and will leverage the power of the platform to host the experiment for free. **Participant’s raw data files (containing identifiers) is automatically stored in a private OSF repository. The preprocessing and analysis scripts, as well as the anonymized data, will be available directly on GitHub, ensuring the transparency and reproducibility of all the analysis steps.**

Participants will be presented with a consent form followed by demographic questions (gender, education level, age, and ethnicity). **Although these variables are directly not analyzed in the current study, they will be used to provide to provide a detailed and thorough description of the sample and maximizing data reusability.**. Participants will then be administered the PCS and the Illusion Game task (IG) in a counterbalanced order.

@@

Participants will be asked to put on their headphones and await further auditory instructions. The PCS procedure starts with a recorded introduction explaining that a series of tests will be applied to evaluate how experiences can be created through imagination. This will be followed by 10 suggestions in a fixed order [see @lush2021], such as "now extend your arms ahead of you, with palms facing each other, hands about a foot apart" and "as you sit comfortably in your chair with your eyes closed, a picture of two balls will be displayed on the computer screen". Once the 10 suggestions are completed, participants will be asked to rate their subjective experiences and response to each suggestion on a 6-points Likert scale. Phenomenological control will be indexed by averaging the scores from the 10 scales.

@@

The task is an adaptation of the one used in @makowski2023 to make it shorter and more reliable, in which participants must make perceptual judgments (e.g., "which red line is the longer") as quickly and accurately as possible. It includes 3 illusion types, namely Ebbinghaus, Müller-Lyer, and Vertical-Horizontal. The procedure encompasses 2 sets of 80 trials for each illusion type. Each set will include, in a random order, the 3 blocks of illusion types, in which trials are separated by a fixation cross, temporally (uniformly sampled duration of 500 - 1000s) and spatially jittered (around the centre of the screen in a radius of a 1 cm) to attenuate its potential usefulness as a reference point. After each illusion type block, a score is presented (computed as a scaled Inverse Efficiency Score) as a gamification mechanism to increase motivation to perform to the best of one's abilities. **To mitigate for the potential variability in the speed/accuracy trade-off, the instructions emphasize with equal weight to be fast and to avoid errors**.

@@

Visual illusion sensitivity will be measured as the average error rate in the incongruent condition, separately for the 3 illusion types. Although the error rate is arguably a crude score, which does not take into account the effect of varying illusion strength, the interaction with task difficulty and the possible adjustments in response strategy (speed-accuracy trade off), it is also the most simple and easy to reproduce, hence its usage as our primary outcome for the current preregistration.

The two sets of 3 illusion blocks will be separated by 2 short questionnaires acting as a break, namely the IPIP-6 [@sibley2011mini], measuring 6 personality traits with 24 analogue scales items, and the PID-5 [@krueger2011], measuring 5 maladaptive personality traits with 25 Likert scales items. **These questionnaires are included as a way of providing a break between the two cognitively taxing blocks and maintain paradigmatic consistency with previous studies [@makowski2023]**.

@@

The PCS will contain several manipulation check indices to identify problematic participants. Participants should not answer "no balls were presented" when queried about the colour they observed on the screen following the negative visual hallucination suggestion, and should execute at least 5 space presses upon instructions to do so.

Illusion Game outliers will be flagged based on their RT distributions, following the same procedure as in [@makowski2023]. If the RT is collapsed to the left (i.e., has \> 1/3 of ultra-fast responses - typically \< 200 ms) in the first block, the entire participant will be discarded (suggesting that they did not properly do the text), but if only the second block is bad, then only the second block will be discarded (as the illusion sensitivity can still be estimated, albeit with less precision). In addition, the removal of individual trials will also be performed [RT \< 200 ms or \> 3 SD; following @theriault2024].

>>>>

Visual Illusions are an interesting type of stimuli highlighting the ease with which our phenomenological conscious experience can become dissociated from physical reality. Their robust and reliable effect makes them useful stimuli to explore how perception is constructed and shaped, and several theoretical models have been put forth to explain how they work. **In particular, illusions have been reframed using a predictive coding account of perception [@notredame2014] in which the brain optimally combines, using some flavour of Bayesian inference, perceptual inputs with prior knowledge to make sense of ambiguous environments [@friston2010].**

@@

However, the exact nature of this interindividual variability and its potential origin remains unclear. The somewhat mixed evidence in the literature regarding its generalizability and strength could be related to the variety of the paradigms used and the type of processes being mobilised [@makowski2021]. Indeed, traditional methods frequently focus on participant's experience by prompting them to assess the difference between two identical targets, estimate the target's physical properties, or adjust the targets to match a reference stimulus [@todorovic2020]. Relying on metacognitive judgments about one's subjective experiences adds an additional layer to the measure that might not be desired when attempting to measure illusion **sensitivity**. Moreover, paradigms often face challenges in diversifying the illusory effects (i.e., using multiple stimuli to experimentally manipulate the strength of the illusion) and the illusion types (i.e., using various illusions, such as Müller-Lyer, Ebbinghaus, Delboeuf which might rely on a different admixture of mechanisms), hindering the potential of obtaining a comprehensive, valid, and reliable measure of illusion sensitivity.

@@

To encourage the empirical exploration of our ability and tendency to alter our phenomenological experience and further accelerate investigations away from the hypnotic context, @lush2021 adapted the Sussex-Waterloo Scale of Hypnotisability [SWASH, @lush2018] by removing all its references to hypnosis, to measure trait phenomenological control. **This newly developed phenomenological control scale (PCS) consists of 10 imaginative suggestions followed by subjective ratings for each suggestion on a 6-point Likert scale (from 0-5) and has demonstrated validity in online experiments [@lush2022].**

@@

The experiment’s setup follows of the born-open principle [@deleeuw2023]. The online experiment, implemented entirely using JsPsych [@de2015jspsych], has its code stored on GitHub and will leverage the power of the platform to host the experiment for free. Participant’s raw data files (containing identifiers) **are** automatically stored in a private OSF repository. The preprocessing and analysis scripts, as well as the anonymized data, will be available directly on GitHub, ensuring the transparency and reproducibility of all the analysis steps.

Participants will be presented with a consent form followed by demographic questions (gender, education level, age, and ethnicity). **Although these variables are not directly analyzed in the current study, they will be used to provide to provide a detailed and thorough description of the sample and maximizing data reusability.** Participants will then be administered the PCS and the Illusion Game task (IG) in a counterbalanced order.

@@

Participants will be asked to put on their headphones and await further auditory instructions. The PCS procedure starts with a recorded introduction explaining that a series of tests will be applied to evaluate how experiences can be created through imagination. This will be followed by 10 suggestions in a fixed order [see @lush2021], such as "now extend your arms ahead of you, with palms facing each other, hands about a foot apart" and "as you sit comfortably in your chair with your eyes closed, a picture of two balls will be displayed on the computer screen". **Once the 10 suggestions are completed, participants will be asked to rate their subjective experiences and response to each suggestion on a 6-points Likert scale (from 0-5).** Phenomenological control will be indexed by averaging the scores from the 10 scales.

@@

The task is an adaptation of the one used in @makowski2023 to make it shorter and more reliable, in which participants must make perceptual judgments (e.g., "which red line is the longer") as quickly and accurately as possible. It includes 3 illusion types, namely Ebbinghaus, Müller-Lyer, and Vertical-Horizontal. The procedure encompasses 2 sets of 80 trials for each illusion type. Each set will include, in a random order, the 3 blocks of illusion types, in which trials are separated by a fixation cross, temporally (uniformly sampled duration of 500 - 1000s) and spatially jittered (around the centre of the screen in a radius of a 1 cm) to attenuate its potential usefulness as a reference point. After each illusion type block, a score is presented (computed as a scaled Inverse Efficiency Score) as a gamification mechanism to increase motivation to perform to the best of one's abilities. To mitigate for the potential variability in the speed/accuracy trade-off, the instructions emphasize with equal weight to be fast and to avoid errors.

@@

Visual illusion sensitivity will be measured as the average error rate in the incongruent condition, separately for the 3 illusion types. Although the error rate is arguably a crude score, which does not take into account the effect of varying illusion strength, the interaction with task difficulty and the possible adjustments in response strategy (speed-accuracy trade off), it is also the most simple and easy to reproduce, hence its usage as our primary outcome for the current **registered report**.

The two sets of 3 illusion blocks will be separated by 2 short questionnaires acting as a break, namely the IPIP-6 [@sibley2011mini], measuring 6 personality traits with 24 analogue scales items, and the PID-5 [@krueger2011], measuring 5 maladaptive personality traits with 25 Likert scales items. These questionnaires are included as a way of providing a break between the two cognitively taxing blocks and maintain paradigmatic consistency with previous studies [@makowski2023].

@@

The PCS will contain several manipulation check indices to identify problematic participants. **The phenomenological control task consists of various auditory and visual exercises. In one such exercise, participants receive the following instruction: “Open your eyes. You will see only two balls on the screen…just two balls”. However, three differently coloured balls are actually displayed. If participants select the option "no balls were shown", it indicates they failed to pay attention to both the auditory instructions and the visual stimuli. In another exercise, participants are asked to press the spacebar six times. If they press it fewer than five times within the allotted time, it suggests a lack of attentiveness to the auditory instructions. Participants will be excluded if they fail at least one of these checks. Participants will be excluded if they fail at least one of the attention checks.**

Illusion Game outliers will be flagged based on their RT distributions, following the same procedure as in [@makowski2023]. **If the RT is collapsed to the left (i.e., has \> 1/3 of ultra-fast responses - typically \< 200 ms) in the first set, the entire participant will be discarded (suggesting that they did not properly do the task), but if only the second set is bad, then only the second set will be discarded (as the illusion sensitivity can still be estimated, albeit with less precision).** In addition, the removal of individual trials will also be performed [RT \< 200 ms or \> 3 SD; following @theriault2024].
